Hydras \& Co.

### 2024-01-22 doc/ordinal-chapter.tex
Small corrections in Chapter 3

\begin{itemize}
\item The ordinals are listed in a strictly increasing order. 
\item Dots : ``$\ldots$'' stand for  infinite sequences of ordinals, not shown for lack of space. For instance, the ordinal $42$ is not shown in the first line, but it exists, somewhere between $17$ and $\omega$.
\item Each ordinal printed in black is the immediate successor of another ordinal. We call it a \emph{successor} ordinal. For instance, $12$ is the successor of $11$, and $\omega^4+1$ the successor of $\omega^4$.
\item Ordinals (displayed in red)  that  follow immediately dots are called \emph{limit ordinals}. With respect to the order induced by this sequence, any limit ordinal $\alpha$ is the least upper bound of  the set $\mathbb{O}_\alpha$ of all ordinals strictly less than $\alpha$.
\item
For instance $\omega$ is the least upper bound of the set of all finite ordinals (in the first line). It is also the first limit ordinal, and the first infinite ordinal, in the sense that 
the set $\mathbb{O}_\omega$ is infinite.
\item The ordinal $\epsilon_0$ is the first number which is equal to its own exponential of base $\omega$. It plays an important role in proof theory, and is particularly studied in chapters~\ref{chap:T1} to \ref{chap:alpha-large}.
\item Any ordinal is  either the ordinal \textcolor{blue}{$0$},
a successor ordinal, or a \textcolor{red}{limit ordinal}.
\end{itemize}

@@

@@

Having the same order type is an equivalence relation between well-ordered sets. Ordinal numbers (in short: \emph{ordinals})  are descriptions (\emph{names}) of the equivalence classes.
For instance, the order type of $(\mathbb{N},<)$ is associated with the ordinal called  $\omega$, and the order we considered on 
the disjoint union of $\mathbb{N}$ and itself is named $\omega+\omega$.

@@

We cannot cite all the literature published on ordinals since Cantor's book
\cite{cantorbook}, and 
leave it to the reader to explore the bibliography.

The introduction of Jos\'e Grimm's report~\cite{grimm:hal-00911710} contains a nice presentation of the main properties of ordinals.

@@

Let $\alpha$ be some (countable) ordinal; 
in \coq{} terms, we call \emph{ordinal notation for $\alpha$} a structure composed 
of:
\begin{itemize}
\item A data type $A$ for representing all ordinals strictly below $\alpha$,
\item A well founded order $<$ on $A$, 
\item A correct function for comparing two ordinals. Note  that the reflexive closure of $<$ is thus a \emph{total order}.
\end{itemize}

@@

The \texttt{Comparable} class, contributed by Jérémy Damour and Théo Zimmermann, allows us to apply generic lemmas and tactics about decidable strict orders.
The correctness of the comparison function is expressed through Stdlib's type 
\texttt{Datatypes.CompareSpec} as specialized by \texttt{Datatypes.CompSpec}.

@@

\inputsnippets{ON_plus/ltWf, ON_plus/OnPlus}

\subsection{The ordinal \texorpdfstring{$\omega+\omega$}{omega + omega}}

The ordinal $\omega+\omega$ (also known as $\omega\times 2$) may be represented as the concatenation 
of two copies of $\omega$ (Figure~\ref{fig:omega-plus-omega}).
It is also represented by the two first lines of Figure~\ref{fig:ordinal-sequence}.

@@

We can define this notation in \coq{} as an instance of \texttt{ON\_plus}.

@@

\inputsnippets{ON_Omega2/plusDef, ON_Omega2/plusExamples}

>>>>

\begin{itemize}
\item The ordinals are listed in a strictly increasing order. 
\item Dots : ``$\ldots$'' stand for  infinite sequences of ordinals, not shown for lack of space. For instance, the ordinal $42$ is not shown in the first line, but it exists, somewhere between $17$ and $\omega$.
\item Each ordinal printed in black is the immediate successor of another ordinal. We call it a \emph{successor} ordinal. For instance, $12$ is the successor of $11$, and $\omega^4+1$ the successor of $\omega^4$.
\item Ordinals (displayed in red)  that  follow immediately dots are called \emph{limit ordinals}. With respect to the order induced by this sequence, any limit ordinal $\alpha$ is the least upper bound of  the set $\mathbb{O}_\alpha$ of all ordinals strictly less than $\alpha$.
For instance,$\omega$ is the least upper bound of the set of all finite ordinals (in the first line). It is also the first limit ordinal, and the first \emph{infinite ordinal number}, in the sense that 
the set $\mathbb{O}_\omega$ is infinite.
\item The ordinal $\epsilon_0$ is the first number which is equal to its own exponential of base $\omega$. It plays an important role in proof theory, and is particularly studied in chapters~\ref{chap:T1} to \ref{chap:alpha-large}.
\item Any ordinal is  either the ordinal \textcolor{blue}{$0$},
a successor ordinal, or a \textcolor{red}{limit ordinal}.
\end{itemize}

@@

We cannot cite all the literature published on ordinals since Cantor's book
\cite{cantorbook}, and 
leave it to the reader to explore the bibliography.
The introduction of Jos\'e Grimm's report~\cite{grimm:hal-00911710} contains also a nice presentation of the main properties of ordinals.

For simplicity's sake, we will only give the definitions which are useful for understanding our \coq development.

@@

Having the same order type is an equivalence relation between well-ordered sets. Ordinal numbers (in short: \emph{ordinals})  are descriptions (\emph{names}) of the associated equivalence classes.
For instance, the order type of $(\mathbb{N},<)$ is associated with the ordinal called  $\omega$, and the order we considered on 
the disjoint union of two copies of $\mathbb{N}$ is associated with  $\omega \times 2$ (a.k.a. $\omega+\omega$).

@@

@@

Let $\alpha$ be some (countable) ordinal; 
we call \emph{ordinal notation for $\alpha$} any structure composed 
of:
\begin{itemize}
\item A data type $A$ for representing all ordinals strictly below $\alpha$,
\item A well founded order $<$ on $A$, 
\item A correct function for comparing two ordinals. Note  that the reflexive closure of $<$ is thus a \emph{total order}.
\end{itemize}

@@

The \texttt{Comparable} class, contributed by Jérémy Damour and Théo Zimmermann, allows us to apply generic lemmas and tactics about decidable strict orders.
The correctness of the comparison function is expressed through Stdlib's type 
\texttt{Datatypes.CompareSpec} and predicate \texttt{Datatypes.CompSpec}.

@@

\inputsnippets{ON_plus/ltWf}

\inputsnippets{ON_plus/OnPlus}

\subsection{Example: The ordinal \texorpdfstring{$\omega+\omega$}{omega + omega}}

The ordinal $\omega+\omega$ (also known as $\omega\times 2$) may be represented as the concatenation 
of two copies of $\omega$ (Figure~\ref{fig:omega-plus-omega}).
It is also represented by the two first lines of Figure~\ref{fig:ordinal-sequence}.
We define this notation in \coq{} as an instance of \texttt{ON\_plus}.

@@

@@

\inputsnippets{ON_Omega2/plusDef}

\inputsnippets{ON_Omega2/plusExamples}
